Mining Interaction Log Data in a Creative Arts Practice using Transition Matrices

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
final updates from Michael and added DOIs

In this note we will show how an archive of collaborative performance
logs can be subjected to a quantitative analysis based on
transition probabilities between gestural states. We will demonstrate
that two measures on transition matrices, flux and entropy, show
significant differences between different kinds of performances, and
give insight into the internal structure of these performances.

Our archive of interaction data-logs has been accumulated to support a
long-term, practice-led, research program into simple iPad interfaces
for group music-making. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from these similarities, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

@@

\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 95    &      &          &         \\
%Record size & 306MB     &         &        &   \\
Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
Participants& 353   & 2    & 4        & 9        \\
Flux &   & 0.02239 & 0.1445 & 0.3246\\
Entropy &   & 0.7683 & 3.342 & 4.442\\          
\hline
\end{tabular}
\caption{
  Representative statistics from the corpus of musical touch-screen
  interaction sessions described in this note.  Each record in the corpus
  consists of low level touch-data recorded in CSV format. 
  Many of the participants were
  present in multiple sessions, but this data was not tracked.\label{corpus-table}}
\end{table}

\section{Macro Gestures and Transition Matrices}

A useful way of modelling user interfaces is to consider their
theoretical configurations as state vectors and to deploy the methods
of matrix algebra to describe transitions between these states during
an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions and manually code
interaction states to produce a transition model. This method has been
applied to usability analysis of many applications outside of the creative arts, for example~\cite{Kannampallil:2007fp}
studied resource-planning software for
security applications and used the transition
probability matrix of these state-sequences to draw
conclusions about the nature of interactions with that software. As an example from the creative
arts, protocols obtained from a computer interface for ``live coding''
computer music have been  coded as states and analysed using
transition matrices~\cite{Swift:2014tya} in both the textual (editing) and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
Because it has been
established that musical performances on touch-screens can contain 
vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}, and that
gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
 can be used identify a sequence of such states for each performer.
If a performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then, over the course of a session, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
process.
If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix (TM), $P$, which characterises
this process by setting each entry, $p_{ij}$, to be the number of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

@@

While the interaction sessions recorded in our corpus were performed
by many different participants in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences, and transition matrices. While
these matrices can be used to visualise the structure of individual
performances, as has been done in ~\cite{Swift:2014tya}, we also wish to extract quantitative
measures from these TMs to compare the large number of sessions in our
corpus.
To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

@@

In the rest of this note we will use these matrix measures to answer
two questions. First, do the transition matrix measures allow us to
distinguish between different performance contexts, types and
instrumental setups? Secondly, do these measures allow us further
insight into the internal structure of performances?

@@

The distribution of flux and entropy with respect to session {\em contexts} (concert,
rehearsal, recording),
{\em types} (improvisation or composition) and {\em instrumentation} (iPads-only, or iPads with acoustic
percussion) are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metadata could
significantly predict the entropy and flux values.

@@

These tests show that the different interaction styles that might be
present in the various session contexts, types and instrumentations in our corpus
appear to be descriminated  by values of the flux and entropy of their transition
matrices.

@@

It appears from our data-mining results that both flux and entropy can be used to
differentiate between session styles and types, but can they be
used to understand some of the internal structure of sessions? Since Aristotle, a
classical model for understanding temporal art-forms is the three-phase
``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
investigate how this structure fits our corpus, we have divided the
gestural-sequences of each session into three equal sections and have calculated the
transition matrices given by each section.

Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable across the three sections than  are
improvisations. This stability of compositions is to be
expected since performers use the same gestures in each rehearsal and
concert of written scores, however it may be that different
kinds of compositions could result in different structural
differences. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the concert performance.

The gestural change in improvisation can be further explored by
dividing the sessions by session context. Figure
\ref{fig:context-section-flux-entropy} shows this division for
improvisations only. In rehearsals and studies the distribution of
flux and entropy over sections has a similar shape, both measures drop
in the ending section of studies and after the beginning section in
rehearsals. In concerts however, flux drops after the beginning
while entropy drops substantially in the ending section. This
indicates a more constrained performance style in this closing
section, where performers transition between fewer gestures in the
final section of concert performances.

@@

While the interaction sessions in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this data collection, different ensembles and
activities have been emphasised and some of these differences can be
seen as clusters of performances having similar fluxes and entropies in the
two plots. For instance, the earliest series of rehearsals and
concerts, from early 2013 until July 2014, 
had fluxes that were bounded between approximately 0.05 and 0.17.
Just before and after July 2014, fluxes were bounded between 0.1 and 0.26. Finally,
between early and mid 2015, fluxes were bounded between 0.08 and 0.32. Within these 
blocks of performances, distinct clusters can be identified. Clusters in the entropy plots
sometimes show similar trends to the fluxes and sometimes different. The higher
upper bounds of fluxes close to July 2015 relative to those a year earlier may indicate
the development of a more ``fluctuating'' style of performance. In the same period, the bound
of entropies remained more similar indicating that the total space of available gestures continued
to be explored in a relatively stable way.

@@

In this note we have applied transition matrices
to a data corpus of 95 collaborative musical performances using touch screens.
We have applied two simple matrix
measures, flux and entropy, to the data and have shown that
 both of these measures appear to 
have discriminatory power. Not only do these measures
differentiate between different types of sessions, and the internal
structure of these sessions, but they provide qualitative insight into the
performances. For improvised performances, it turns out that our results showed that concerts
had the lowest values of both flux and entropy, while recording sessions had
the highest. This may be due to the increased pressure on performers to take fewer risks
when improvising in front of an audience at a concert and indicates that perhaps rehearsals should 
emphasise the development of such risk-taking. 
Our results for the three phases of performance (beginning, middle, end)
revealed more variation in
improvisations than compositions. As a creative
 response to this finding, 
new works might be
composed that specifically explore and emphasise different gestural sequences across the three phases.

>>>>

In this note we show how an archive of collaborative performance logs
can be quantitatively analysed using estimated transition
probabilities between gestural states. We will demonstrate that two
measures on transition matrices, flux and entropy, show significant
differences between different kinds of performances, and yield insight
into the internal structure of these performances.

Our archive of interaction data-logs has been accumulated to support a
long-term, practice-led, research program into simple iPad interfaces
for group music-making. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where participants used our custom-built, touch-screen apps on Apple
iPad devices. Although several different apps have been included in
our corpus, they shared a common mode of free-form touch interaction
where tapping produced short sounds and swirling or swiping produced
long sounds. Apart from these similarities, the apps featured a
variety of sound palettes, different kinds of visual feedback, and
different kinds of networked interactions (between iPads in an
ensemble and with a central server).

@@

\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 95    &      &          &         \\
%Record size & 306MB     &         &        &   \\
Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
Participants& 353   & 2    & 4        & 9        \\
Flux &   & 0.02239 & 0.1445 & 0.3246\\
Entropy &   & 0.7683 & 3.342 & 4.442\\          
\hline
\end{tabular}
\caption{
  Representative statistics from the corpus of musical touch-screen
  interaction sessions.  Each record in the corpus
  consists of low level touch-data recorded in CSV format. 
  Many of the participants were
  present in multiple sessions, but this data was not tracked.\label{corpus-table}}
\end{table}

\break 
\section{Macro Gestures and Transition Matrices}

A useful way of modelling user interfaces is to consider their
theoretical configurations as state vectors and to describe transitions between these states during
an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions and manually code
interaction states to produce a transition model. This method has been
applied to usability analysis of many applications outside of the creative arts, for example~\cite{Kannampallil:2007fp}
studied resource-planning software for
security applications and used the transition
probability matrix of these state-sequences to draw
conclusions about the nature of interactions with that software. As an example from the creative
arts, protocols obtained from a computer interface for ``live coding''
computer music have been  coded as states and analysed using
transition matrices~\cite{Swift:2014tya} in both the textual (editing) and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
Because it has been
established that musical performances on touch-screens define  
vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}, and that
gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
 can be used identify a sequence of such states for each performer.
If a performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then, over the course of a session, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be modelled as a Markov
process.
If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix (TM), $P$, which characterises
this process by setting each entry, $p_{ij}$, to be the proportion of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

@@

While the interaction sessions recorded in our corpus were performed
by many different participants in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences, and transition matrices. While
these matrices can be used to visualise the structure of individual
performances, as has been done in ~\cite{Swift:2014tya}, we also wish to extract quantitative
measures from these TMs to compare the large number of sessions in our
corpus.
To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two consecutive measurements.

@@

We use these matrix measures to answer
two questions. First, do the transition matrix measures allow us to
distinguish between different performance contexts, types and
instrumental setups? Secondly, do these measures allow us further
insight into the internal structure of performances?

@@

The distributions of flux and entropy with respect to session {\em contexts} (concert,
rehearsal, recording),
{\em types} (improvisation or composition) and {\em instrumentation} (iPads-only, or iPads with acoustic
percussion) are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed to find whether the metadata could 
successfully predict the entropy and flux values.

@@

These tests suggest that the different interaction styles that might be
present in the various session contexts, types and instrumentations in our corpus
appear to be discriminated  by values of the flux and entropy of their transition
matrices.

@@

It appears from our data-mining results that each of flux and entropy can be used to
differentiate between session styles and types, but can they be
used to understand some of the internal structure of sessions? Since Aristotle, a
classical model for understanding temporal art-forms is the three-phase
``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
investigate how this structure fits our corpus, we divided the
gestural-sequences of each session into three equal sections and calculated the
transition matrices given by each section.

Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable across the three sections than  are
improvisations. This stability of compositions is to be
expected since performers use the same gestures in each rehearsal and
concert of written scores, however it may be that different
kinds of compositions could result in different structural
patterns. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the concert performance.

The gestural change in improvisation can be further explored by
dividing the sessions by session context. Figure
\ref{fig:context-section-flux-entropy} shows this division for
improvisations only. In rehearsals and recordings the distributions of
flux and entropy over sections has a similar shape, both measures dropping
in the ending section of recordings and after the beginning section in
rehearsals. In concerts however, flux drops after the beginning
while entropy drops substantially in the ending section. This
indicates a more constrained performance style in this closing
section, where performers transition between fewer gestures in the
final section of concert performances.

@@

While the interaction sessions in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this data collection, different ensembles and
activities have been emphasised and some of these differences can be
seen as clusters of performances having similar fluxes and entropies in the
two panels. For instance, the earliest series of rehearsals and
concerts, from early 2013 until July 2014, 
had fluxes that were between approximately 0.05 and 0.17.
Just before and after July 2014, fluxes were between 0.1 and 0.26. Finally,
between early and mid 2015, fluxes were between 0.08 and 0.32. Within these 
blocks of performances, distinct clusters can be identified. The higher
upper bounds of fluxes close to July 2015 relative to those a year earlier may indicate
the development of a more ``fluctuating'' style of performance over time. In the same period, the bound
of entropies remained more consistent indicating that the total space of available gestures continued
to be explored in a relatively stable way.

@@

In this note we estimated transition matrices 
for a data corpus of 95 collaborative musical performances using touch screens.
We have applied two simple matrix
measures, flux and entropy, to the data and have shown that
 both of these measures appear to 
have discriminatory power. Not only do these measures
differentiate between different types of sessions, and the internal
structure of these sessions, but they provide qualitative insight into the
performances. For improvised performances, our results showed that concerts
had the lowest values of both flux and entropy, while recording sessions had
the highest. This may be due to the increased pressure on performers to take fewer risks
when improvising in front of an audience at a concert and indicates that perhaps rehearsals should 
emphasise the development of such risk-taking. 
Our results for the three phases of performance (beginning, middle, end)
revealed more variation in
improvisations than compositions. As a creative
 response to this finding, 
new works might be
composed that specifically explore and emphasise different gestural sequences across the three phases.
